Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-23 Multiresolution_framework/ltx/manuscript.tex
added block diagram for the EM algorithm
minor edits

Both the E and the M steps of the EM algorithm find increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For state-space model the E step corresponds to the smoothing problem which can be solved using the Rauch Tung Streibel (RTS) smoother \cite{RAUCH1965}. For the M-step the lower bound on the likelihood function should be constructed, the steps needed
to compute the required quantities are very dependent on the particular application which is described in detail for our problem. In summary, to solve the joint state and parameter estimation for the state space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure
  \subsection{E-step}
  As mentioned earlier the E-step for the state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the state of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$. Following \cite{Gibsona2005}, an additional backwards recursion is included (step 6 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required to construct the lower bound. Note that the following quantities from the forward iteration should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates and covariance matrices, and also the final value of the Kalman gain in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
\begin{equation}
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \Theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \Theta)\right], 
\end{equation}
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix...)  where $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ is the joint probability distribution of the states and the observations.

>>>>

Both the E and the M steps of the EM algorithm find increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For state-space model the E step corresponds to the smoothing problem which can be solved using the Rauch Tung Streibel (RTS) smoother \cite{RAUCH1965}. For the M-step the lower bound on the likelihood function should be constructed, the steps needed
to compute the required quantities are very dependent on the particular application which is described in detail for our problem. In summary, to solve the joint state and parameter estimation for the state space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/EMBlockDiagram.pdf}
\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}
\label{fig:EMBlockDiagram}
\end{figure}
  \subsection{E-step}
  As mentioned earlier the E-step for the state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the state of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$. Following \cite{Gibsona2005}, an additional backwards recursion is included (step 6 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required to construct the lower bound. Note that the following quantities from the forward iteration should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates and covariance matrices, and also the final value of the Kalman gain in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
\begin{equation}
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \Theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \Theta)\right], 
\end{equation}
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix...)  where $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ is the joint probability distribution of the states and the observations.
